---
title: "Module 2 — Intro to R and Quarto"
format: html
editor: source
execute:
  warning: false
  message: false
---

This is your workspace for the Module 2 worksheet. The full instructions live
on the lesson page — keep it open alongside this notebook. You will build up
this notebook one part at a time: fill in the empty chunk under each heading,
press **Run Cell** to see what happens, and write a sentence of your own
between the chunks. Render with **Preview** whenever the worksheet tells you to.

Always load the packages you need in a setup chunk at the top. Run this first,
every time you open the notebook:

```{r setup}
library(tidyverse)
library(here)
```

## Part 1 — Run a chunk, write a sentence, render

Let's start simple. This chunk does a basic calculation in R and stores the
result in a variable. R prints the result automatically when you run it.

```{r}
# Add two numbers together
ten <- 3 + 7
ten
```

R can also work with text (called **strings**). The `c()` function combines
values into a **vector** — a list of things of the same type.

```{r}
# Make a vector of numbers and calculate the average
my_numbers <- c(12, 18, 25, 33, 47)
mean(my_numbers)
```

**My sentence:** In the first chunk, I assigned `3 + 7` to a variable called
`ten`, and R printed `10`. In the second chunk, I made a vector of five numbers
and `mean()` calculated their average (which is `27`).

## Part 2 — Read the data and look at it

Here we read in the instructional dataset files using `read_csv()`. This function
comes from the `readr` package (part of the tidyverse) and turns a CSV file into
a **data frame** — R's version of a spreadsheet with rows and columns.

```{r}
# Read the sample IDs file — this tells us which samples belong to which participant
samples <- read_csv(here("data", "instructional_dataset", "00_sample_ids_UKZN_workshop_2023.csv"))

# Read participant metadata — age, smoking status, education, etc.
metadata <- read_csv(here("data", "instructional_dataset", "01_participant_metadata_UKZN_workshop_2023.csv"))

# Read clinical measurements — Nugent score, CRP, pH
clinical <- read_csv(here("data", "instructional_dataset", "02_visit_clinical_measurements_UKZN_workshop_2023.csv"))
```

`glimpse()` gives us a quick summary — it shows the number of rows, the column
names, their data types (e.g. `<chr>` = text, `<dbl>` = decimal number), and a
peek at the first few values.

```{r}
# Look at the structure of each dataset
glimpse(samples)
glimpse(metadata)
glimpse(clinical)

# head() shows the first 6 rows — handy for a quick sanity check
head(samples)
```

`summary()` gives us basic statistics for each column (min, max, mean, etc. for
numbers; counts for text columns).

```{r}
summary(metadata)
summary(clinical)
```

## Part 3 — Pull out columns and summarise them

```{r}
# your code here
```

## Part 4 — Build something with Pi

```{r}
# your code here
```

## Capstone — make it your own

```{r}
# your code here
```

## Before you go

Pick one step today that confused you. Ask Pi to explain it, then push back:
*"why did you use X instead of Y?"* Notice when Pi sounds confident versus when
it is guessing.
