---
title: "Does diet affect treatment? — Yogurt and the vaginal microbiome"
format: html
editor: source
execute:
  warning: false
  message: false
---

> **Question:** If you eat yogurt, does your vaginal microbiome respond
> differently to antibiotic treatment compared to keeping your usual diet?

The study had 51 participants. Half ate yogurt daily, the other half kept
their usual diet. Everyone got antibiotics, and researchers measured
bacterial levels at **baseline** (before antibiotics) and **after
antibiotic** treatment.

The bacteria measured are:

- **Total bacteria** — all bacteria together
- ***L. crispatus*** — a "good" bacterium linked to vaginal health
- ***L. iners*** — another common vaginal bacterium

```{r setup}
library(tidyverse)
library(here)
library(ggpubr)

# Read in the yogurt study data
sample_ids <- read_csv(here("data/group_a_yogurt/00_sample_ids_yogurt.csv"))
metadata   <- read_csv(here("data/group_a_yogurt/01_participant_metadata_yogurt.csv"))
qpcr       <- read_csv(here("data/group_a_yogurt/02_qpcr_results_yogurt.csv"))
```

## Join the data together

We connect the tables by `sample_id` so each bacterial measurement also has
the participant's diet group (`arm`) and time point.

```{r join}
# Join qPCR results with sample info (time_point, arm)
bacteria_data <- qpcr |>
  left_join(sample_ids, by = "sample_id") |>
  mutate(time_point = fct_relevel(time_point, "baseline", "after_antibiotic"))

# Take a peek
bacteria_data
```

## Boxplot: bacterial levels by diet group

Each plot below compares the **yogurt** group (pink) to the **unchanged
diet** group (blue) — separately at baseline and after antibiotics. The
p-value tells us whether the difference between groups is larger than what
we'd expect by chance.

```{r bacteria-boxplots}
#| fig-width: 8
#| fig-height: 6

# Reshape the data to long format for faceting
bacteria_long <- bacteria_data |>
  pivot_longer(
    cols = c(qpcr_bacteria, qpcr_crispatus, qpcr_iners),
    names_to = "bacteria_type",
    values_to = "count"
  ) |>
  mutate(
    # Clean up labels for the plot
    bacteria_type = case_match(
      bacteria_type,
      "qpcr_bacteria" ~ "Total bacteria",
      "qpcr_crispatus" ~ "L. crispatus",
      "qpcr_iners" ~ "L. iners"
    ),
    bacteria_type = fct_inorder(bacteria_type)
  )

# Boxplot faceted by time point and bacteria type
ggplot(bacteria_long, aes(x = arm, y = count, fill = arm)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.7) +
  geom_jitter(width = 0.15, alpha = 0.5, size = 1.5) +
  facet_grid(bacteria_type ~ time_point, scales = "free_y") +
  scale_fill_manual(values = c("unchanged_diet" = "#4A90D9", "yogurt" = "#E86A92")) +
  stat_compare_means(
    method = "wilcox.test",
    label.x = 1.3,
    size = 3
  ) +
  labs(
    title = "Does diet affect the microbiome's response to antibiotics?",
    subtitle = "Comparing yogurt vs unchanged diet at baseline and after antibiotic treatment",
    x = "Diet group",
    y = "Bacterial count (qPCR)",
    fill = "Diet group"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold")
  )
```

### What to look for

- **Baseline column** — Are the groups already different before the
  antibiotic? If the p-value is small (< 0.05), the groups weren't well
  matched at the start.
- **After antibiotic column** — Does the diet group make a difference after
  treatment? A small p-value here suggests eating (or not eating) yogurt
  changed how the microbiome responded to antibiotics.

---

## Paired look: how each person changed

This plot shows each participant's bacterial levels **before** and **after**
antibiotics, with a line connecting the same person. This lets us see:

- Did *L. crispatus* go up or down for each person?
- Did the yogurt group recover differently than the unchanged diet group?

```{r paired-plots}
#| fig-width: 9
#| fig-height: 7

bacteria_long |>
  ggplot(aes(x = time_point, y = count)) +
  geom_line(aes(group = pid), alpha = 0.25, linewidth = 0.4) +
  geom_point(aes(color = arm), size = 2, alpha = 0.7) +
  facet_grid(bacteria_type ~ arm, scales = "free_y") +
  scale_color_manual(values = c("unchanged_diet" = "#4A90D9", "yogurt" = "#E86A92")) +
  stat_compare_means(
    method = "wilcox.test",
    paired = TRUE,
    comparisons = list(c("baseline", "after_antibiotic")),
    size = 3
  ) +
  labs(
    title = "How did each person's bacteria levels change?",
    subtitle = "Each line is one participant, from baseline → after antibiotics",
    x = "Time point",
    y = "Bacterial count (qPCR)",
    color = "Diet group"
  ) +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    strip.text = element_text(face = "bold", size = 10),
    plot.title = element_text(face = "bold")
  )
```

### What to look for

- The p-value in each panel tells you if the **change** from baseline to
  after antibiotics was significant within that group.
- Compare the direction of change between the **yogurt** and **unchanged
  diet** columns. If lines mostly go in opposite directions (or one group
  changes a lot while the other stays flat), that suggests diet affected the
  treatment response.