---
title: "Group B — Menstruation Study: Data Dictionary"
subtitle: "Exploring the menstrual health dataset"
format:
  html:
    toc: true
    code-fold: true
    embed-resources: true
---

```{r}
#| include: false
library(tidyverse)
```

## Overview

This dataset comes from a **menstrual health study** (Group B). It contains
data from 27 participants across four linked CSV files:

| # | File | What it contains |
|---|------|------------------|
| 1 | `00_sample_ids_period.csv` | Links participants to their samples across time points |
| 2 | `01_participant_metadata_period.csv` | Demographics and health info about each participant |
| 3 | `02_luminex_period.csv` | Cytokine/chemokine concentrations measured via Luminex |
| 4 | `03_flow_cytometry_period.csv` | Immune cell counts measured via flow cytometry |

The study has two main groupings:

- **Arm** (*birth_control* vs *no_birth_control*) — whether the participant
  uses hormonal birth control.
- **Time point** (*onset*, *end_bleeding*, *week_prior*, *week_post*) — when
  the sample was collected relative to the participant's menstrual period.

---

## File 1: Sample IDs (`00_sample_ids_period.csv`)

**What a row represents:** One biological sample collected from a participant
at a specific time point.

```{r}
sample_ids <- read_csv("../../data/group_b_menstruation/00_sample_ids_period.csv")
glimpse(sample_ids)
```

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | **Participant ID** — a unique identifier for each person in the study (e.g., `pid_01`, `pid_12`). Links to `01_participant_metadata_period.csv`. |
| `time_point` | character | **Menstrual cycle time point** when the sample was collected. One of: `onset` (start of period), `end_bleeding` (end of bleeding), `week_prior` (one week before onset), `week_post` (one week after end of bleeding). |
| `arm` | character | **Study arm** — whether the participant uses hormonal birth control (`birth_control`) or not (`no_birth_control`). This is redundant with the metadata file. |
| `sample_id` | character | **Unique sample identifier** (e.g., `SAMP001`). This is the key used to link to the Luminex and flow cytometry data. Each participant may contribute multiple samples (one per time point). |

---

## File 2: Participant Metadata (`01_participant_metadata_period.csv`)

**What a row represents:** One participant enrolled in the study.

```{r}
metadata <- read_csv("../../data/group_b_menstruation/01_participant_metadata_period.csv")
glimpse(metadata)
```

| Column | Type | Meaning |
|--------|------|---------|
| `pid` | character | **Participant ID** — matches the `pid` in the sample IDs file. Each participant appears exactly once here. |
| `arm` | character | **Study arm** — `birth_control` (uses hormonal birth control) or `no_birth_control` (does not use hormonal birth control). This is the main grouping variable for the study. |
| `age` | double | **Age of the participant** in years. Ranges from 26 to 41 in this dataset. |
| `pcos_status` | character | **PCOS diagnosis status** — whether the participant has Polycystic Ovary Syndrome. Values: `pcos` (diagnosed) or `no disease` (no diagnosis). |
| `period_product` | character | **Menstrual product used** — `tampon`, `pad`, or `menstrual_cup`. |
| `sex` | character | **Sex assigned at birth** — all participants are `female` in this dataset. |

---

## File 3: Luminex Results (`02_luminex_period.csv`)

**What a row represents:** The concentration of one cytokine or chemokine
measured in one biological sample.

```{r}
luminex <- read_csv("../../data/group_b_menstruation/02_luminex_period.csv")
glimpse(luminex)
```

This is a **long/tidy** format: each row is one analyte measurement for one
sample. To see it as a wide table, you would pivot by `cytokine`.

```{r}
# How many unique cytokines are measured?
luminex |> 
  distinct(cytokine) |> 
  pull(cytokine)
```

```{r}
# How many samples and how are values censored?
luminex |> 
  count(limits)
```

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | **Sample identifier** — matches `sample_id` in the sample IDs file. Links each measurement to a participant and time point. |
| `cytokine` | character | **Name of the analyte measured**. The 8 cytokines/chemokines in this panel are: `IL-1a`, `IL-1b`, `IL-6`, `TNFa`, `MIG` (CXCL9), `IFNg`, `IP-10` (CXCL10), and `MIP-3a` (CCL20). These are immune signaling proteins involved in inflammation and immune regulation. |
| `conc` | double | **Concentration** of the cytokine in the sample (units are relative Luminex fluorescence / standard curve units). Numeric value. |
| `limits` | character | **Censoring status** — tells you if the concentration was reliably detectable. One of: `not_censored` (measured successfully), `below_detection_limit` (too low to detect), `above_detect_limit` (too high — above the assay's upper limit). These values need special handling in analysis (e.g., imputation or threshold modeling). |

---

## File 4: Flow Cytometry (`03_flow_cytometry_period.csv`)

**What a row represents:** The immune cell counts from one biological sample,
broken down by several cell types measured via flow cytometry.

```{r}
flow <- read_csv("../../data/group_b_menstruation/03_flow_cytometry_period.csv")
glimpse(flow)
```

```{r}
# Summary of the cell count columns
flow |> 
  select(-sample_id) |> 
  summary()
```

| Column | Type | Meaning |
|--------|------|---------|
| `sample_id` | character | **Sample identifier** — matches `sample_id` in the sample IDs file. |
| `live_cd19_negative` | double | **Total live cells that are CD19−** (all live cells that are not B cells). This is a gating/parent population. |
| `cd45_negative` | double | **CD45− cells** — cells that do not express CD45 (the leukocyte common antigen). These are typically non-immune cells (e.g., epithelial, stromal, or red blood cells). |
| `cd45_positive` | double | **CD45+ cells** — white blood cells / leukocytes. These are immune cells that express CD45. |
| `neutrophils` | double | **Neutrophil count** — a type of granulocyte and the most abundant white blood cell. Key in innate immunity and inflammation. |
| `non_neutrophils` | double | **CD45+ cells that are not neutrophils** — all other white blood cells (lymphocytes, monocytes, eosinophils, basophils, etc.). |
| `cd3_negative` | double | **CD3− cells** (within the non-neutrophil population) — immune cells that are not T cells (e.g., B cells, NK cells, monocytes). |
| `cd3_positive` | double | **CD3+ cells** — T lymphocytes (T cells). CD3 is a marker expressed by all T cells. |
| `cd4_t_cells` | double | **CD4+ T cells** — helper T cells (also called CD4+ T lymphocytes). These coordinate immune responses. |
| `cd8_t_cells` | double | **CD8+ T cells** — cytotoxic T cells (also called CD8+ T lymphocytes). These kill infected or abnormal cells. |

**Gating hierarchy (how the cell types relate to each other):**

```
All live CD19− cells
├── CD45− (non-immune cells)
└── CD45+ (leukocytes)
    ├── Neutrophils
    └── Non-neutrophils
        ├── CD3− (non-T cells: B cells, NK cells, monocytes, etc.)
        └── CD3+ (T cells)
            ├── CD4+ T cells (helper T cells)
            └── CD8+ T cells (cytotoxic T cells)
```

Note that `cd4_t_cells` + `cd8_t_cells` do **not** necessarily equal
`cd3_positive` — there are other T cell subsets (like gamma-delta T cells)
that are CD3+ but neither CD4+ nor CD8+.

---

## Relationships Between Files

The files can be joined together using these key columns:

```
01_participant_metadata     00_sample_ids              02_luminex
┌─────────────┐            ┌─────────────┐            ┌─────────────┐
│ pid         │◄──────┐    │ pid         │            │ sample_id   │
│ arm         │       │    │ time_point  │            │ cytokine    │
│ age         │       │    │ arm         │            │ conc        │
│ pcos_status │       │    │ sample_id   │──┐         │ limits      │
│ period_product│      │    └─────────────┘  │         └─────────────┘
│ sex         │       │                      │
└─────────────┘       │                      │
                      │                      │
                      │         03_flow_cytometry
                      │         ┌─────────────┐
                      │         │ sample_id   │◄─┘
                      │         │ live_cd19…  │
                      └─────────│ cd45_…      │
                                │ neutrophils │
                                │ …           │
                                └─────────────┘
```

To build a complete analysis dataset joining all files:

```{r}
#| code-fold: true
# Join sample IDs with participant metadata
full_data <- sample_ids |> 
  left_join(metadata, by = "pid", suffix = c("", "_meta")) |> 
  # Drop the redundant arm column from sample_ids (we keep the one from metadata)
  select(-arm) |> 
  rename(arm = arm_meta)

# Join with Luminex (long format)
full_data_luminex <- full_data |> 
  left_join(luminex, by = "sample_id")

# Join with flow cytometry
full_data_flow <- full_data |> 
  left_join(flow, by = "sample_id")

# Preview the combined datasets
glimpse(full_data_luminex)
```

```{r}
glimpse(full_data_flow)
```